---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# tcrdistR

C++-accelerated TCR distance calculations for T-cell receptor repertoire
analysis. Computes pairwise TCRdist distances incorporating V-region and CDR3
sequence comparisons using BLOSUM62-derived substitution matrices. Targets
feature parity with Python
[tcrdist3](https://github.com/kmayerb/tcrdist3).

## Installation

Install the development version from GitHub:

```r
# install.packages("devtools")
devtools::install_github("shihanli92/tcrdistR")
```

## Quick Start

```{r quick-start, eval=FALSE}
library(tcrdistR)
data(dash)

# Build a TCRrep object (deduplicates identical clones per subject)
rep <- TCRrep(dash, organism = "mouse", compute_distances = TRUE)
rep
#> TCRrep with 1888 clones (mouse, AB chains)
#>   paired_dist: 1888 x 1888

# Pairwise distance heatmap
plot_tcrdist_heatmap(rep@paired_dist[1:40, 1:40])

# Kernel PCA
pca <- compute_tcrdist_kernel_pca(
  rep@clone_df, rep@organism, n_components = 50L
)
plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = rep@clone_df$epitope,
  title = "Kernel PCA",
  point_size = 1.5
)

# UMAP (uses TCRdist KNN neighbors directly)
umap <- compute_tcrdist_umap(rep@clone_df, rep@organism, seed = 42)
plot_tcr_scatter(
  umap$embeddings,
  color_by = rep@clone_df$epitope,
  axis_label_prefix = "UMAP",
  point_size = 1.5
)
```

## Documentation

Full reference and vignettes at
**<https://shihanli92.github.io/tcrdistR/>**.

- `vignette("tcrdistR-getting-started")` -- Installation and core functions
- `vignette("tcrdistR-tcrrep-workflow")` -- End-to-end analysis with the
  TCRrep object
- `vignette("tcrdistR-advanced")` -- Clumping, meta-clonotypes, clustering,
  DB matching
- `vignette("tcrdistR-visualization")` -- Visualization guide

## License

MIT
