---
title: "Advanced TCR Repertoire Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced TCR Repertoire Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

This vignette demonstrates the advanced analysis features in tcrdistR:
hierarchical clustering, neighborhood statistical tests, clumping detection,
meta-clonotype discovery, database matching, and more.

We use a small inline dataset of mouse TCRs from different subjects responding
to different epitopes:

```{r setup, eval=FALSE}
library(tcrdistR)

tcrs <- data.frame(
  va    = c("TRAV7-3*01",  "TRAV6D-6*01", "TRAV6D-6*01",
            "TRAV6-4*01",  "TRAV6-4*01",  "TRAV7-3*01",
            "TRAV6D-6*01", "TRAV6-4*01",  "TRAV7-3*01",
            "TRAV6D-6*01", "TRAV6-4*01",  "TRAV7-3*01",
            "TRAV6D-6*01", "TRAV6-4*01",  "TRAV7-3*01"),
  cdr3a = c("CAVSLDSNYQLIW",  "CALGDRATGGNNKLTF", "CALGSNTGYQNFYF",
            "CALAPSNTNKVVF",  "CALVPSNTNKVVF",    "CAVSLDSNYQLIW",
            "CALGDRATGGNNKLTF","CALAPSNTNKVVF",    "CAVSLDSNYQLIW",
            "CALGSNTGYQNFYF", "CALVPSNTNKVVF",    "CAVSLDSNYQLIW",
            "CALGDRATGGNNKLTF","CALAPSNTNKVVF",    "CAVSLDSNYQLIW"),
  vb    = c("TRBV13-1*01", "TRBV29*01",   "TRBV29*01",
            "TRBV2*01",    "TRBV29*01",   "TRBV13-1*01",
            "TRBV29*01",   "TRBV2*01",    "TRBV13-1*01",
            "TRBV29*01",   "TRBV29*01",   "TRBV13-1*01",
            "TRBV29*01",   "TRBV2*01",    "TRBV13-1*01"),
  cdr3b = c("CASSDFDWGGDAETLYF", "CASSPDRGEVFF",    "CASTGGGAPLF",
            "CASSQDPGDYEQYF",   "CASSLGGENTLYF",   "CASSDFDWGGDAETLYF",
            "CASSPDRGEVFF",      "CASSQDPGDYEQYF",  "CASSDFDWGGDAETLYF",
            "CASTGGGAPLF",       "CASSLGGENTLYF",    "CASSDFDWGGDAETLYF",
            "CASSPDRGEVFF",      "CASSQDPGDYEQYF",  "CASSDFDWGGDAETLYF"),
  epitope = c("PA", "PA", "PA", "PA", "PA",
              "NP", "NP", "NP", "NP", "NP",
              "M1", "M1", "M1", "M1", "M1"),
  subject = c("S1", "S1", "S2", "S2", "S3",
              "S1", "S2", "S3", "S1", "S2",
              "S3", "S1", "S2", "S3", "S1"),
  stringsAsFactors = FALSE
)
```

## Hierarchical Clustering

Cluster TCRs by distance using hierarchical agglomerative clustering:

```{r hclust, eval=FALSE}
hc <- tcrdist_hclust(tcrs, organism = "mouse", method = "average")

# hc contains:
# - hc$hclust: an hclust object (pass to plot(), cutree(), etc.)
# - hc$dist_matrix: the underlying distance matrix
# - hc$indices: row indices used (relevant if subsampled)

plot(hc$hclust, main = "TCR Hierarchical Clustering")
```

Cut the dendrogram into a fixed number of clusters:

```{r cluster, eval=FALSE}
clusters <- cluster_tcrs(tcrs, organism = "mouse", k = 3)
table(clusters)

# Or cut by height threshold
clusters_h <- cluster_tcrs(tcrs, organism = "mouse", h = 100)
```

## Neighborhood Statistical Tests

Test whether a categorical variable (e.g., epitope specificity) is enriched
in TCR neighborhoods:

```{r nhood-test, eval=FALSE}
# Fisher's exact test for binary variables
result <- neighborhood_test(
  tcrs, organism = "mouse",
  variable = ifelse(tcrs$epitope == "PA", "PA", "other"),
  radius = 100, test = "fisher"
)

# Each row reports the test result for one TCR's neighborhood
head(result[, c("cdr3a", "cdr3b", "odds_ratio", "p_value", "p_adjusted")])

# Chi-squared test for multi-level variables
result_chi <- neighborhood_test(
  tcrs, organism = "mouse",
  variable = tcrs$epitope,
  radius = 100, test = "chisq"
)
```

## Meta-Clonotype Discovery

Identify quasi-public TCR motifs shared across multiple subjects:

```{r meta-clonotypes, eval=FALSE}
meta <- find_meta_clonotypes(
  tcrs, organism = "mouse",
  radius = 100,
  min_nsubject = 2L,
  subject_col = "subject"
)

# Returns a data.frame with center TCR info, radius, neighbor count,
# number of subjects, and neighbor indices
print(meta[, c("cdr3a", "cdr3b", "radius", "K_neighbors", "nsubject")])
```

Summarize the composition of a discovered meta-clonotype:

```{r summarize-meta, eval=FALSE}
if (nrow(meta) > 0) {
  indices <- as.integer(strsplit(meta$neighbor_indices[1], ",")[[1]])
  summary <- summarize_meta_clonotype(tcrs, meta$center_index[1], indices)
  summary$n_members
  summary$va_usage
  summary$vb_usage
}
```

## Database Matching

Match query TCRs against the built-in literature epitope databases:

```{r db-match, eval=FALSE}
# Paired-chain matching with background-corrected p-values
results <- match_tcrs_to_db(tcrs, organism = "mouse")
significant <- results[results$pvalue_adj < 0.05, ]

# Single-chain exact CDR3 matching (no background model needed)
hits <- strict_single_chain_match_tcrs_to_db(tcrs, organism = "mouse")
hits$alpha_matches  # matches on CDR3-alpha
hits$beta_matches   # matches on CDR3-beta
```

## CD8 Scoring

Score TCRs for CD8 phenotype using a logistic regression model:

```{r cd8, eval=FALSE}
# Requires human TCRs
human_tcrs <- data.frame(
  va    = c("TRAV1-1*01", "TRAV1-2*01"),
  cdr3a = c("CAVRDSSYKLIF", "CAVKDSSYKLIF"),
  vb    = c("TRBV5-1*01", "TRBV6-1*01"),
  cdr3b = c("CASSIRSSYEQYF", "CASSIKSSYEQYF"),
  stringsAsFactors = FALSE
)

scores <- make_cd8_score_table_column(human_tcrs)
# Returns a numeric vector of CD8 probability scores
```

## Kernel PCA

Embed TCRs in lower-dimensional space via kernel PCA on the distance matrix:

```{r kpca, eval=FALSE}
pca <- compute_tcrdist_kernel_pca(tcrs, organism = "mouse", n_components = 10)

# pca$embeddings: N x n_components matrix
# pca$eigenvalues: eigenvalues
# pca$indices: row indices used

# Use with plot_tcr_scatter for visualization
plot_tcr_scatter(
  pca$embeddings[, 1], pca$embeddings[, 2],
  color = tcrs$epitope[pca$indices],
  xlab = "KPC1", ylab = "KPC2"
)
```

## Fuzzy Diversity

Compute TCR-aware diversity that accounts for sequence similarity:

```{r fuzzy-div, eval=FALSE}
fd <- tcr_fuzzy_diversity(tcrs, organism = "mouse", threshold = 50)
fd$fuzzy_diversity     # diversity accounting for similar TCRs
fd$standard_diversity  # standard Simpson's for comparison
```

## Distance-Based Joins

Fuzzy-join two TCR datasets by distance threshold:

```{r join, eval=FALSE}
left <- tcrs[1:5, ]
right <- tcrs[6:15, ]

# Inner join: only matched pairs within radius
joined <- tcrdist_join(left, right, organism = "mouse",
                       radius = 100, max_n = 3)

# Left join: all left rows, NAs for unmatched
joined_left <- tcrdist_join(left, right, organism = "mouse",
                            radius = 100, type = "left")
```
