---
title: "Visualization Guide"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Visualization Guide}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
has_viz <- requireNamespace("ggplot2", quietly = TRUE)
has_logo <- has_viz && requireNamespace("ggseqlogo", quietly = TRUE)
```

## Setup

tcrdistR provides plotting functions built on ggplot2. All return ggplot
objects that can be further customized.

```{r setup}
library(tcrdistR)
data(dash)

# Use 100 PA-specific TCRs for fast plots
pa <- dash[dash$epitope == "PA", ]
pa_sub <- pa[1:100, ]
```

## Distance Heatmap

Visualize the pairwise distance matrix as a heatmap. Note that
`plot_tcrdist_heatmap()` takes a precomputed distance matrix, not a TCR
data.frame:

```{r heatmap, eval=has_viz, fig.height=5}
library(ggplot2)

dist_mat <- tcrdist_matrix(pa_sub, organism = "mouse")
plot_tcrdist_heatmap(dist_mat)
```

With hierarchical clustering reordering (the default) and custom labels:

```{r heatmap-labels, eval=has_viz, fig.height=5}
# Use a small subset with unique labels for readability
small_mat <- dist_mat[1:20, 1:20]
plot_tcrdist_heatmap(
  small_mat,
  labels = paste0(pa_sub$cdr3b[1:20], " (", seq_len(20), ")"),
  cluster = TRUE,
  title = "PA-specific TCR Distances"
)
```

## Dendrogram

Plot a hierarchical clustering dendrogram. Unlike the heatmap,
`plot_tcrdist_dendrogram()` takes a TCR data.frame and computes distances
internally:

```{r dendrogram, eval=has_viz, fig.height=4}
plot_tcrdist_dendrogram(pa_sub, organism = "mouse")
```

Color by a grouping variable:

```{r dendrogram-colored, eval=has_viz, fig.height=4}
# Mix epitopes for a more interesting dendrogram
mixed <- dash[dash$epitope %in% c("PA", "NP"), ]
mixed_sub <- mixed[sample(nrow(mixed), 80), ]

plot_tcrdist_dendrogram(
  mixed_sub, organism = "mouse",
  color_by = mixed_sub$epitope,
  title = "PA vs NP TCRs"
)
```

## Distance Distribution

Histogram and density of pairwise distances. Like the heatmap, this takes a
precomputed distance matrix:

```{r dist-distrib, eval=has_viz}
plot_distance_distribution(dist_mat)
```

With custom bin width:

```{r dist-distrib-bins, eval=has_viz}
plot_distance_distribution(dist_mat, binwidth = 10,
                           title = "PA TCR distance distribution")
```

## PCA Scatter Plot

2D scatter plot from kernel PCA or any dimensionality reduction.
`plot_tcr_scatter()` takes a N x 2 coordinate matrix:

```{r scatter, eval=has_viz, fig.height=5}
pca <- compute_tcrdist_kernel_pca(
  mixed_sub, organism = "mouse", n_components = 5L
)

plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = mixed_sub$epitope,
  title = "Kernel PCA: PA vs NP",
  point_size = 2
)
```

Color by a continuous variable (e.g., third principal component):

```{r scatter-continuous, eval=has_viz, fig.height=5}
plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = pca$embeddings[, 3],
  title = "Colored by KPC3",
  point_size = 2,
  legend_title = "KPC3"
)
```

## Gene Usage

Horizontal bar chart of V-gene frequencies. `plot_gene_usage()` takes a
data.frame and a column name:

```{r gene-usage, eval=has_viz}
plot_gene_usage(pa, "va", title = "V-alpha usage (PA)")
plot_gene_usage(pa, "vb", title = "V-beta usage (PA)")
```

## CDR3 Sequence Logos

Display amino acid frequency at each position as a sequence logo. The
logo construction aligns sequences using BLOSUM62-optimal gap placement:

```{r logo, eval=has_logo, fig.height=3}
plot_cdr3_logo(pa_sub$cdr3b, chain = "beta", method = "bits",
               title = "CDR3-beta logo (PA)")
```

Frequency ("prob") mode:

```{r logo-prob, eval=has_logo, fig.height=3}
plot_cdr3_logo(pa_sub$cdr3a, chain = "alpha", method = "prob",
               title = "CDR3-alpha logo (PA)")
```

## Combining Plots

Use the patchwork package to arrange multiple plots:

```{r patchwork, eval=has_viz && requireNamespace("patchwork", quietly = TRUE), fig.width=8, fig.height=6}
library(patchwork)

p1 <- plot_tcrdist_heatmap(dist_mat[1:30, 1:30], title = "Heatmap")
p2 <- plot_distance_distribution(dist_mat, title = "Distribution")
p3 <- plot_gene_usage(pa, "va", title = "V-alpha")
p4 <- plot_gene_usage(pa, "vb", title = "V-beta")

(p1 | p2) / (p3 | p4) +
  plot_annotation(title = "PA Repertoire Overview")
```
