---
title: "Getting Started with tcrdistR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tcrdistR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**tcrdistR** computes pairwise distances between T-cell receptors (TCRs) using
the TCRdist metric. Each TCR is represented by its V-gene and CDR3 amino acid
sequence for both alpha and beta chains. The distance incorporates V-region
similarity (via BLOSUM62-derived substitution matrices) and CDR3 sequence
alignment with variable gap positioning matching tcrdist3.

All distance computations are implemented in C++ for high performance.

## Installation

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("shihanli92/tcrdistR")
```

## The DASH Dataset

tcrdistR ships with the **DASH** dataset (Dash et al., 2017) --- 1924 paired
alpha-beta mouse TCRs responding to 7 viral epitopes, collected from 78
subjects. We use it throughout these vignettes.

```{r load-data}
library(tcrdistR)
data(dash)

dim(dash)
colnames(dash)
table(dash$epitope)
```

The required columns for distance computation are `va`, `cdr3a`, `vb`, and
`cdr3b`. Additional columns like `epitope` and `subject` are carried along for
downstream analysis.

## Reading TCR Data from Files

tcrdistR can read TCR data from multiple common formats. All readers produce
a data.frame with standardized column names (`va`, `cdr3a`, `vb`, `cdr3b`).

```{r read-data, eval=FALSE}
# Auto-detect format from column naming conventions
tcrs <- read_tcr_table("my_data.tsv")

# Format-specific readers
tcrs <- read_10x("filtered_contig_annotations.csv")
tcrs <- read_airr("airr_rearrangements.tsv")
tcrs <- read_adaptive("immunoseq_export.tsv")
```

## Computing the Distance Matrix

The core function `tcrdist_matrix()` computes a dense N x N pairwise distance
matrix. We use a subset of 50 PA-specific TCRs for speed:

```{r dist-matrix}
pa <- dash[dash$epitope == "PA", ]
pa_sub <- pa[1:50, ]

dist_mat <- tcrdist_matrix(pa_sub, organism = "mouse")
dim(dist_mat)
dist_mat[1:5, 1:5]
```

Each entry is an integer TCRdist value. Identical TCRs have distance 0;
typical distances range from 0 to ~400.

## Sparse Distances

For large datasets, storing the full N x N matrix is impractical. Use
`tcrdist_sparse()` to only store distances below a threshold:

```{r sparse}
sparse_mat <- tcrdist_sparse(pa_sub, organism = "mouse", threshold = 100)
class(sparse_mat)
```

The result is a `Matrix::dgCMatrix`. Non-stored entries represent distances
above the threshold.

## Rectangular Distances

Compute distances between two different sets of TCRs (query vs reference):

```{r rect}
query <- pa_sub[1:5, ]
reference <- pa_sub[6:50, ]
rect_mat <- tcrdist_rect(query, reference, organism = "mouse")
dim(rect_mat)
```

## K-Nearest Neighbors

Find the K closest TCRs for each input. This computes distances and extracts
neighbors in a single pass without materializing the full N x N matrix:

```{r knn}
knn <- tcrdist_knn(pa_sub, organism = "mouse", K = 5L)
dim(knn$knn_indices)    # N x K
dim(knn$knn_distances)  # N x K

# Nearest neighbor of TCR 1
knn$knn_indices[1, ]
knn$knn_distances[1, ]
```

## Radius-Based Neighbors

Find all neighbors within a distance threshold:

```{r radius}
neighbors <- tcrdist_radius_neighbors(pa_sub, organism = "mouse", radius = 50)
length(neighbors)  # one entry per TCR

# First TCR's neighbors
neighbors[[1]]$indices
neighbors[[1]]$distances
```

## Per-Component Distances

All distance functions accept a `components` parameter to compute distances
from specific TCR regions. This is useful for comparing CDR3-only distances,
V-region-only contributions, or single-chain distances.

```{r components}
# CDR3 only (both chains, no V-region)
d_cdr3 <- tcrdist_matrix(pa_sub, "mouse", components = "cdr3")

# V-region only (CDR1 + CDR2 + CDR2.5)
d_v <- tcrdist_matrix(pa_sub, "mouse", components = "v_region")

# Components are additive: v_region + cdr3 = all
d_all <- tcrdist_matrix(pa_sub, "mouse")
stopifnot(all(abs((d_v + d_cdr3) - d_all) < 1e-10))

# Single chain
d_alpha <- tcrdist_matrix(pa_sub, "mouse", components = "alpha")
d_beta  <- tcrdist_matrix(pa_sub, "mouse", components = "beta")
stopifnot(all(abs((d_alpha + d_beta) - d_all) < 1e-10))

# Custom combination
d_custom <- tcrdist_matrix(pa_sub, "mouse", components = c("cdr3a", "vb"))
```

Available presets: `"all"` (default), `"cdr3"`, `"v_region"`, `"alpha"`,
`"beta"`. Individual terms: `"va"`, `"cdr3a"`, `"vb"`, `"cdr3b"`.

## Single-Chain Mode

Many datasets contain only beta-chain data (e.g., from bulk sequencing). All
distance functions automatically detect single-chain input and compute
distances for the available chain only:
```{r single-chain}
# Beta-only data (no alpha columns)
beta_only <- pa_sub[, c("vb", "cdr3b")]
d_beta_only <- tcrdist_matrix(beta_only, "mouse")
dim(d_beta_only)

# Result matches components="beta" on paired data
d_beta_paired <- tcrdist_matrix(pa_sub, "mouse", components = "beta")
stopifnot(all(abs(d_beta_only - d_beta_paired) < 1e-10))

# Works with all wrappers: sparse, rect, knn, radius_neighbors
knn_beta <- tcrdist_knn(beta_only, "mouse", K = 3L)
```

## Standardizing TCR Data

Use `as_tcr_df()` to convert data.frames from common tools (scRepertoire,
scirpy, dandelion, tcrdist3) to tcrdistR's canonical column names:

```{r as-tcr-df, eval=FALSE}
# Auto-detect format
tcrs <- as_tcr_df(my_data)

# Explicit format
tcrs <- as_tcr_df(my_data, format = "screpertoire")

# Custom column mapping
tcrs <- as_tcr_df(my_data, col_map = c(va = "V_alpha", cdr3a = "CDR3_alpha",
                                        vb = "V_beta",  cdr3b = "CDR3_beta"))
```

## Single-Pair Distances

For comparing individual CDR3 sequences:

```{r single-pair}
# Weighted CDR3 distance (uses BLOSUM62-derived BSD4 matrix)
weighted_cdr3_distance("CAVSLDSNYQLIW", "CALGDRATGGNNKLTF")

# Simple Hamming distance (counts mismatches at each position)
hamming_distance("CASSI", "CASSK")
```

## Diversity Metrics

Repertoire diversity functions work on clone count vectors:

```{r diversity}
# Clone counts for PA-specific TCRs
pa_counts <- pa$count

# Generalized diversity (order 2 = inverse Simpson)
div <- tcr_diversity(pa_counts, order = 2)
div$entropy
div$effective_number

# Species richness
tcr_richness(pa_counts)

# Clonality (1 - normalized Shannon entropy)
tcr_clonality(pa_counts)
```

## The TCRrep Object

For more structured workflows, wrap your data in a `TCRrep` S4 object.
By default, `TCRrep()` deduplicates identical clones (matching tcrdist3
behavior), grouping by chain columns plus `subject` if present, and summing
clone counts:

```{r tcrrep}
rep <- TCRrep(pa_sub, organism = "mouse")
nrow(pa_sub)        # before dedup
nrow(rep@clone_df)  # after dedup (within-subject duplicates merged)
rep
```

You can control deduplication with the `deduplicate` argument:

```{r tcrrep-dedup}
# Chain columns only (collapse across subjects)
rep_strict <- TCRrep(pa_sub, organism = "mouse", deduplicate = character(0))
nrow(rep_strict@clone_df)

# No deduplication
rep_raw <- TCRrep(pa_sub, organism = "mouse", deduplicate = FALSE)
nrow(rep_raw@clone_df)
```

## Next Steps

- `vignette("tcrdistR-advanced")` --- Clustering, neighborhood tests,
  meta-clonotypes, database matching, kernel PCA, fuzzy diversity
- `vignette("tcrdistR-visualization")` --- Heatmaps, dendrograms, CDR3 logos,
  scatter plots
