---
title: "TCRrep Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCRrep Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
has_viz <- requireNamespace("ggplot2", quietly = TRUE)
has_logo <- has_viz && requireNamespace("ggseqlogo", quietly = TRUE)
has_patch <- has_viz && requireNamespace("patchwork", quietly = TRUE)
```

## Introduction

The `TCRrep` S4 object is the central data structure in tcrdistR. It holds
the clonotype table, organism, distance matrices, and downstream results in a
single object. This vignette walks through a complete analysis using TCRrep
as the backbone.

## Creating a TCRrep

Load the DASH dataset and construct a TCRrep. By default, identical clones
within the same subject are deduplicated and their counts summed:

```{r create}
library(tcrdistR)
data(dash)

rep <- TCRrep(dash, organism = "mouse", compute_distances = TRUE)
rep
```

The `compute_distances = TRUE` flag computes the full pairwise distance matrix
at construction time and stores it in the `paired_dist` slot:

```{r dist-slot}
dim(rep@paired_dist)
rep@paired_dist[1:5, 1:5]
```

## Distance Heatmap

With distances already computed, pass the `paired_dist` slot directly to the
heatmap. We use a subset for readability:

```{r heatmap, eval=has_viz, fig.height=5}
library(ggplot2)

# First 50 clones
idx <- 1:50
plot_tcrdist_heatmap(rep@paired_dist[idx, idx],
                     title = "TCRdist heatmap (first 50 clones)")
```

## Distance Distribution

```{r dist-distrib, eval=has_viz}
plot_distance_distribution(rep@paired_dist, title = "Pairwise distance distribution")
```

## Gene Usage

Gene usage plots take the clone data.frame and a column name. Access the
clonotype table via `rep@clone_df`:

```{r gene-usage, eval=has_viz}
plot_gene_usage(rep@clone_df, "va", title = "V-alpha usage")
plot_gene_usage(rep@clone_df, "vb", title = "V-beta usage")
```

## Dendrogram

The dendrogram function takes a data.frame and organism --- extract from the
TCRrep slots. We subsample for a cleaner plot:

```{r dendro, eval=has_viz, fig.height=5}
sub_idx <- sample(nrow(rep@clone_df), 100)
plot_tcrdist_dendrogram(
  rep@clone_df[sub_idx, ],
  organism = rep@organism,
  color_by = rep@clone_df$epitope[sub_idx],
  title = "TCR dendrogram colored by epitope"
)
```

## Kernel PCA and Scatter Plots

Compute a kernel PCA embedding from the TCRrep:

```{r kpca}
pca <- compute_tcrdist_kernel_pca(
  rep@clone_df, organism = rep@organism,
  n_components = 10L, method = "eigen"
)
dim(pca$embeddings)
```

Visualize the first two components, colored by epitope:

```{r scatter, eval=has_viz, fig.height=5}
plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = rep@clone_df$epitope,
  title = "Kernel PCA of DASH TCRs",
  point_size = 1.5
)
```

## Clustering

Cluster TCRs and add cluster assignments to the clone table:

```{r cluster}
rep@clone_df$cluster <- cluster_tcrs(
  rep@clone_df, organism = rep@organism, k = 7
)
table(rep@clone_df$cluster)
```

Visualize clusters in PCA space:

```{r scatter-cluster, eval=has_viz, fig.height=5}
plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = as.factor(rep@clone_df$cluster),
  title = "Clusters in PCA space",
  point_size = 1.5,
  legend_title = "Cluster"
)
```

## CDR3 Sequence Logos by Epitope

Extract CDR3 sequences for a specific epitope and plot a logo:

```{r logo-pa, eval=has_logo, fig.height=3}
pa_idx <- rep@clone_df$epitope == "PA"
plot_cdr3_logo(
  rep@clone_df$cdr3b[pa_idx],
  chain = "beta", method = "bits",
  title = "CDR3-beta logo (PA epitope)"
)
```

```{r logo-np, eval=has_logo, fig.height=3}
np_idx <- rep@clone_df$epitope == "NP"
plot_cdr3_logo(
  rep@clone_df$cdr3b[np_idx],
  chain = "beta", method = "bits",
  title = "CDR3-beta logo (NP epitope)"
)
```

## Neighborhood Test

Test whether epitope labels are enriched in TCR neighborhoods:

```{r nhood}
nhood <- neighborhood_test(
  rep@clone_df, organism = rep@organism,
  variable = rep@clone_df$epitope,
  radius = 50, test = "chisq"
)
head(nhood[order(nhood$p_adjusted), c("index", "n_neighbors", "p_value", "p_adjusted")])
```

## Meta-Clonotypes

Find TCR motifs shared across subjects:

```{r meta}
meta <- find_meta_clonotypes(
  rep@clone_df, organism = rep@organism,
  radius = 48, min_nsubject = 2L,
  subject_col = "subject"
)
nrow(meta)
head(meta[, c("cdr3a", "cdr3b", "radius", "K_neighbors", "nsubject")])
```

Store back into the TCRrep:

```{r meta-store}
rep@meta_clonotypes <- meta
```

## Diversity

Compute repertoire diversity using the clone counts stored in the TCRrep:

```{r diversity}
div <- tcr_diversity(rep@clone_df$count, order = 2)
div$effective_number

tcr_clonality(rep@clone_df$count)
```

## Combined Panel

```{r panel, eval=has_patch, fig.width=10, fig.height=8}
library(patchwork)

p1 <- plot_tcrdist_heatmap(rep@paired_dist[1:40, 1:40], title = "Heatmap")
p2 <- plot_distance_distribution(rep@paired_dist, title = "Distribution")
p3 <- plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = rep@clone_df$epitope,
  title = "Kernel PCA", point_size = 1
)
p4 <- plot_gene_usage(rep@clone_df, "vb", title = "V-beta usage")

(p1 | p2) / (p3 | p4) +
  plot_annotation(title = "DASH Repertoire Overview")
```
