---
title: "Advanced TCR Repertoire Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced TCR Repertoire Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

This vignette demonstrates the advanced analysis features in tcrdistR:
hierarchical clustering, neighborhood statistical tests, clumping detection,
meta-clonotype discovery, database matching, kernel PCA, and more.

We use the DASH dataset --- 1924 paired alpha-beta mouse TCRs across 7 viral
epitopes from 78 subjects.

```{r setup}
library(tcrdistR)
data(dash)

# Work with PA-specific TCRs for clustering demos (N = 324)
pa <- dash[dash$epitope == "PA", ]
nrow(pa)
```

## Hierarchical Clustering

Cluster TCRs by distance using hierarchical agglomerative clustering:

```{r hclust}
hc <- tcrdist_hclust(pa, organism = "mouse", method = "average")

# hc contains:
#   hc$hclust      - an hclust object (pass to plot(), cutree(), etc.)
#   hc$dist_matrix - the underlying distance matrix
#   hc$indices     - row indices used (relevant if subsampled)
names(hc)
```

Cut the dendrogram into a fixed number of clusters:

```{r cluster}
clusters <- cluster_tcrs(pa, organism = "mouse", k = 5)
table(clusters)

# Or cut by height threshold
clusters_h <- cluster_tcrs(pa, organism = "mouse", h = 100)
table(clusters_h)
```

## Neighborhood Statistical Tests

Test whether epitope specificity is enriched in TCR neighborhoods. This
computes TCRdist radius-based neighborhoods and runs a statistical test for
each TCR.

```{r nhood-test}
# Use a mixed-epitope subset for a meaningful test
mixed <- dash[dash$epitope %in% c("PA", "NP"), ]
mixed$is_PA <- ifelse(mixed$epitope == "PA", "PA", "other")

result <- neighborhood_test(
  mixed, organism = "mouse",
  variable = mixed$is_PA,
  radius = 50, test = "fisher"
)

head(result[order(result$p_adjusted), c("index", "n_neighbors", "odds_ratio",
                                         "p_value", "p_adjusted")])
```

For multi-level categorical variables, use the chi-squared test:

```{r nhood-chisq, eval=FALSE}
result_chi <- neighborhood_test(
  dash, organism = "mouse",
  variable = dash$epitope,
  radius = 50, test = "chisq"
)
```

## Meta-Clonotype Discovery

Identify quasi-public TCR motifs shared across multiple subjects. A
meta-clonotype is defined by a center TCR and a radius: all TCRs within the
radius from multiple subjects form the meta-clonotype.

```{r meta-clonotypes}
meta <- find_meta_clonotypes(
  pa, organism = "mouse",
  radius = 48,
  min_nsubject = 2L,
  subject_col = "subject"
)

nrow(meta)
head(meta[, c("cdr3a", "cdr3b", "radius", "K_neighbors", "nsubject")])
```

Summarize the composition of a discovered meta-clonotype:

```{r summarize-meta}
if (nrow(meta) > 0) {
  indices <- as.integer(strsplit(meta$neighbor_indices[1], ",")[[1]])
  summary <- summarize_meta_clonotype(pa, meta$center_index[1], indices)
  summary$n_members
  summary$va_usage
  summary$vb_usage
}
```

## Clumping Detection

Test whether TCRs form statistically significant clusters beyond what is
expected by chance. This uses a background resampling model to compute
expected neighbor counts.

The `find_clumping()` function requires nucleotide sequences and J-gene
annotations, which the DASH dataset provides:

```{r clumping, eval=FALSE}
clump <- find_clumping(
  pa, organism = "mouse",
  radii = c(24L, 48L, 72L),
  num_random_samples = 50000L,
  verbose = TRUE
)

# How many TCRs are in significant clumps?
sum(clump$is_clumped)

# Detailed results sorted by significance
head(clump$results_df[, c("cdr3a", "cdr3b", "nbr_radius",
                            "pvalue_adj", "num_nbrs")])
```

## Database Matching

Match query TCRs against the built-in literature epitope database. This
requires background sampling and is available for human TCRs:

```{r db-match, eval=FALSE}
# Paired-chain matching with background-corrected p-values
results <- match_tcrs_to_db(human_tcrs, organism = "human")
significant <- results[results$pvalue_adj < 0.05, ]

# Single-chain exact CDR3 matching (no background model needed)
hits <- strict_single_chain_match_tcrs_to_db(human_tcrs, organism = "human")
hits$alpha_matches
hits$beta_matches
```

## Kernel PCA

Embed TCRs in lower-dimensional space via kernel PCA on the distance matrix.
The default kernel is linear (`1 - D/Dmax`); a Gaussian kernel is also
available.

```{r kpca}
pca <- compute_tcrdist_kernel_pca(
  pa, organism = "mouse",
  n_components = 10L,
  method = "eigen"
)

# Return structure
names(pca)
dim(pca$embeddings)     # N x n_components
length(pca$eigenvalues)

# Top eigenvalues (variance captured)
head(pca$eigenvalues)
```

With Gaussian kernel:

```{r kpca-gaussian}
pca_gauss <- compute_tcrdist_kernel_pca(
  pa, organism = "mouse",
  n_components = 10L,
  kernel = "gaussian",
  gaussian_kernel_sdev = 100
)
head(pca_gauss$eigenvalues)
```

## Fuzzy Diversity

Compute TCR-aware diversity that accounts for sequence similarity. Unlike
standard diversity which treats each unique clonotype as completely distinct,
fuzzy diversity merges clonotypes within a distance threshold:

```{r fuzzy-div}
fd <- tcr_fuzzy_diversity(pa, organism = "mouse", threshold = 50)
fd$fuzzy_diversity     # diversity accounting for similar TCRs
fd$standard_diversity  # standard Simpson's for comparison
```

## Distance-Based Joins

Fuzzy-join two TCR datasets by distance threshold. This finds all pairs of
TCRs (one from each dataset) within a specified radius:

```{r join}
left <- pa[1:20, ]
right <- pa[21:100, ]

# Inner join: only matched pairs within radius
joined <- tcrdist_join(left, right, organism = "mouse",
                       radius = 100, max_n = 3)
dim(joined)
head(joined[, c("cdr3a_x", "cdr3b_x", "cdr3a_y", "cdr3b_y", "tcrdist")])
```

## CD8 Scoring

Score TCRs for CD8 phenotype using a logistic regression model trained on
known CD4/CD8 repertoires. This is only available for human TCRs:

```{r cd8, eval=FALSE}
scores <- make_cd8_score_table_column(human_tcrs)
# Returns a numeric vector of CD8 probability scores
```
