---
title: "Getting Started with tcrdistR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tcrdistR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**tcrdistR** computes pairwise distances between T-cell receptors (TCRs) using
the TCRdist metric. Each TCR is represented by its V-gene and CDR3 amino acid
sequence for both alpha and beta chains. The distance incorporates V-region
similarity (via BLOSUM62-derived substitution matrices) and CDR3 sequence
alignment.

All distance computations are implemented in C++ for high performance. This
vignette walks through the core functionality.

## Installation

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("shihanli92/tcrdistR")
```

## Creating TCR Data

tcrdistR works with data.frames containing at minimum four columns: `va`,
`cdr3a`, `vb`, `cdr3b` (V-gene and CDR3 for alpha and beta chains).

```{r create-data, eval=FALSE}
library(tcrdistR)

tcrs <- data.frame(
  va    = c("TRAV7-3*01", "TRAV6D-6*01", "TRAV6D-6*01",
            "TRAV6-4*01", "TRAV6-4*01"),
  cdr3a = c("CAVSLDSNYQLIW", "CALGDRATGGNNKLTF", "CALGSNTGYQNFYF",
            "CALAPSNTNKVVF", "CALVPSNTNKVVF"),
  vb    = c("TRBV13-1*01", "TRBV29*01", "TRBV29*01",
            "TRBV2*01", "TRBV29*01"),
  cdr3b = c("CASSDFDWGGDAETLYF", "CASSPDRGEVFF", "CASTGGGAPLF",
            "CASSQDPGDYEQYF", "CASSLGGENTLYF"),
  stringsAsFactors = FALSE
)
```

## Reading TCR Data from Files

tcrdistR can read TCR data from multiple common formats:

```{r read-data, eval=FALSE}
# Auto-detect format from column naming conventions
tcrs <- read_tcr_table("my_data.csv")

# Format-specific readers
tcrs <- read_10x("filtered_contig_annotations.csv")
tcrs <- read_airr("airr_rearrangements.tsv")
tcrs <- read_adaptive("immunoseq_export.tsv")
```

## The TCRrep Object

For more structured workflows, wrap your data in a `TCRrep` S4 object:

```{r tcrrep, eval=FALSE}
rep <- TCRrep(tcrs, organism = "mouse", chains = "paired")
rep
```

## Computing the Distance Matrix

The core function `tcrdist_matrix()` computes a dense N x N pairwise distance
matrix:

```{r dist-matrix, eval=FALSE}
dist_mat <- tcrdist_matrix(tcrs, organism = "mouse")
dim(dist_mat)   # 5 x 5
dist_mat[1:3, 1:3]
```

Each entry is an integer TCRdist value. Identical TCRs have distance 0;
typical distances range from 0 to ~400.

## Sparse Distances

For large datasets, storing the full N x N matrix is impractical. Use
`tcrdist_sparse()` to only store distances below a threshold:

```{r sparse, eval=FALSE}
sparse_mat <- tcrdist_sparse(tcrs, organism = "mouse", threshold = 100)
class(sparse_mat)  # dgCMatrix (sparse)
```

The result is a `Matrix::dgCMatrix` where zero entries represent distances
above the threshold (not distance 0 --- actual zeros are stored as well).

## Rectangular Distances

Compute distances between two different sets of TCRs (query vs reference):

```{r rect, eval=FALSE}
query <- tcrs[1:2, ]
reference <- tcrs[3:5, ]
rect_mat <- tcrdist_rect(query, reference, organism = "mouse")
dim(rect_mat)  # 2 x 3
```

## K-Nearest Neighbors

Find the K closest TCRs for each input:

```{r knn, eval=FALSE}
knn <- tcrdist_knn(tcrs, organism = "mouse", K = 3)
knn$knn_indices   # K x N matrix of neighbor indices
knn$knn_distances  # K x N matrix of distances
```

## Radius-Based Neighbors

Find all neighbors within a distance threshold:

```{r radius, eval=FALSE}
neighbors <- tcrdist_radius_neighbors(
  tcrs, organism = "mouse", radius = 50
)
# Returns a list of neighbor indices and distances per TCR
```

## Single-Pair Distances

For comparing individual CDR3 sequences:

```{r single-pair, eval=FALSE}
# Weighted CDR3 distance (uses BLOSUM62-derived BSD4 matrix)
d <- weighted_cdr3_distance("CAVRDSSYKLIF", "CAVKDSSYKLIF")

# Simple Hamming distance (counts mismatches)
h <- hamming_distance("CASSI", "CASSK")  # returns 1
```

## Diversity Metrics

Repertoire diversity functions work on count vectors (no C++ needed):

```{r diversity}
library(tcrdistR)

# Clone counts for 5 clonotypes
counts <- c(100, 50, 30, 15, 5)

# Generalized Simpson's diversity (order 2)
div <- tcr_diversity(counts, order = 2)
div$entropy
div$effective_number

# Species richness (number of unique clonotypes)
tcr_richness(counts)

# Clonality (1 - normalized Shannon entropy)
tcr_clonality(counts)
```

## Next Steps

- `vignette("tcrdistR-advanced")` --- Clumping, meta-clonotypes, clustering,
  database matching, kernel PCA
- `vignette("tcrdistR-visualization")` --- Heatmaps, dendrograms, CDR3 logos,
  scatter plots
