---
title: "Getting Started with tcrdistR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tcrdistR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**tcrdistR** computes pairwise distances between T-cell receptors (TCRs) using
the TCRdist metric. Each TCR is represented by its V-gene and CDR3 amino acid
sequence for both alpha and beta chains. The distance incorporates V-region
similarity (via BLOSUM62-derived substitution matrices) and CDR3 sequence
alignment with variable gap positioning matching tcrdist3.

All distance computations are implemented in C++ for high performance.

## Installation

tcrdistR uses C++ code (via Rcpp) for fast distance computation, so you need
a C++ compiler:

- **macOS**: Install Xcode Command Line Tools by running
  `xcode-select --install` in Terminal.
- **Windows**: Install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
  (match the version to your R version).
- **Linux**: Install `r-base-dev` (Debian/Ubuntu) or `R-devel` (Fedora/RHEL).

Then install from GitHub:

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("shihanli92/tcrdistR")
```

**Optional dependencies** for visualization and UMAP (install any you need):

```{r optional-deps, eval=FALSE}
install.packages(c("ggplot2", "ggseqlogo", "patchwork", "uwot", "igraph"))
```

## The DASH Dataset

tcrdistR ships with the **DASH** dataset (Dash et al., 2017) --- 1924 paired
alpha-beta mouse TCRs responding to 7 viral epitopes, collected from 78
subjects. We use it throughout these vignettes.

```{r load-data}
library(tcrdistR)
data(dash)

dim(dash)
colnames(dash)
table(dash$epitope)
```

The required columns for distance computation are `va`, `cdr3a`, `vb`, and
`cdr3b`. Additional columns like `epitope` and `subject` are carried along for
downstream analysis.

## Loading Your Own Data

If you have a CSV or TSV file from a sequencing core, you can load it
directly. tcrdistR needs four columns for paired-chain analysis: `va`
(V-alpha gene), `cdr3a` (CDR3-alpha amino acid sequence), `vb` (V-beta
gene), and `cdr3b` (CDR3-beta amino acid sequence). If you have only
beta-chain data, you need just `vb` and `cdr3b`.

```{r own-data, eval=FALSE}
# Step 1: Read your file
my_data <- read.csv("my_tcr_data.csv")

# Step 2: Check your columns
colnames(my_data)
# e.g., [1] "V_alpha" "CDR3_alpha" "V_beta" "CDR3_beta" "patient"

# Step 3: Map your column names to tcrdistR names
tcrs <- as_tcr_df(my_data, col_map = c(
  va   = "V_alpha",
  cdr3a = "CDR3_alpha",
  vb   = "V_beta",
  cdr3b = "CDR3_beta"
))

# Step 4: Check the result
head(tcrs[, c("va", "cdr3a", "vb", "cdr3b")])
```

V-gene names should look like `TRAV1-1*01` or `TRBV5-1*01`. If your genes
lack allele suffixes (e.g., `TRAV1-1` instead of `TRAV1-1*01`), set
`normalize_genes = TRUE` in `as_tcr_df()` to append `*01` automatically.

**Beta-only data** (common from bulk sequencing): just provide `vb` and
`cdr3b` columns. All distance functions automatically detect single-chain
input:

```{r beta-only-load, eval=FALSE}
beta_tcrs <- as_tcr_df(my_data, col_map = c(
  vb   = "V_beta",
  cdr3b = "CDR3_beta"
))
d <- tcrdist_matrix(beta_tcrs, "human")  # works with beta-only data
```

## Reading from Common Formats

tcrdistR also has dedicated readers for common pipeline outputs. All
produce a data.frame with standardized column names.

```{r read-data, eval=FALSE}
# Auto-detect format from column naming conventions
tcrs <- read_tcr_table("my_data.tsv")

# 10X Genomics (produces paired alpha-beta data; cells with only one chain
# are dropped with a message showing how many were excluded)
tcrs <- read_10x("filtered_contig_annotations.csv")

# AIRR Community Standard format (paired if both chains present)
tcrs <- read_airr("airr_rearrangements.tsv")

# Adaptive Biotechnologies (beta-chain only)
tcrs <- read_adaptive("immunoseq_export.tsv")
```

## Computing the Distance Matrix

**What is TCRdist?** TCRdist quantifies how similar two TCR sequences are.
It aligns the CDR3 regions of both TCRs, scores each amino acid
match/mismatch using a biochemical similarity matrix (BSD4, derived from
BLOSUM62), adds a penalty for insertions/deletions (gaps), then adds the
distance between the germline-encoded CDR1, CDR2, and CDR2.5 loops (looked
up from the V-gene). Lower distance means more similar TCRs. TCRs with
small distances (roughly < 50) often recognize the same antigen.

The core function `tcrdist_matrix()` computes a dense N x N pairwise distance
matrix. We use a subset of 50 PA-specific TCRs for speed:

```{r dist-matrix}
pa <- dash[dash$epitope == "PA", ]
pa_sub <- pa[1:50, ]

dist_mat <- tcrdist_matrix(pa_sub, organism = "mouse")
dim(dist_mat)
dist_mat[1:5, 1:5]
```

Each entry is an integer TCRdist value. Identical TCRs have distance 0;
typical distances range from 0 to ~400. As a rough guide:

- **0--25**: Very similar; likely recognize the same epitope with similar
  binding mode.
- **25--75**: Moderately similar; may share specificity.
- **75--200**: Distantly related; different CDR3 or V-gene usage.
- **> 200**: Unrelated.

## Sparse Distances

For large datasets, the full N x N matrix becomes impractical --- a 10,000
TCR dataset needs ~800 MB, and 50,000 TCRs would need ~20 GB. Use
`tcrdist_sparse()` to only store distances below a threshold:

```{r sparse}
sparse_mat <- tcrdist_sparse(pa_sub, organism = "mouse", threshold = 100)
class(sparse_mat)
```

The result is a `Matrix::dgCMatrix` (a compressed sparse column matrix).
Non-stored entries represent distances above the threshold. Since most TCR
pairs are unrelated, the sparse matrix is dramatically smaller than the
dense version.

## Rectangular Distances

Compute distances between two different sets of TCRs (query vs reference):

```{r rect}
query <- pa_sub[1:5, ]
reference <- pa_sub[6:50, ]
rect_mat <- tcrdist_rect(query, reference, organism = "mouse")
dim(rect_mat)
```

## K-Nearest Neighbors

Find the K closest TCRs for each input. This computes distances and extracts
neighbors in a single pass without materializing the full N x N matrix:

```{r knn}
knn <- tcrdist_knn(pa_sub, organism = "mouse", K = 5L)
dim(knn$knn_indices)    # N x K
dim(knn$knn_distances)  # N x K

# Nearest neighbor of TCR 1
knn$knn_indices[1, ]
knn$knn_distances[1, ]
```

## Radius-Based Neighbors

Find all neighbors within a distance threshold:

```{r radius}
neighbors <- tcrdist_radius_neighbors(pa_sub, organism = "mouse", radius = 50)
length(neighbors)  # one entry per TCR

# First TCR's neighbors
neighbors[[1]]$indices
neighbors[[1]]$distances
```

## Per-Component Distances

All distance functions accept a `components` parameter to compute distances
from specific TCR regions. This lets you ask focused biological questions:

- **CDR3 only** (`"cdr3"`): captures the unique rearrangement-derived
  binding interface --- the most variable part of the TCR.
- **V-region only** (`"v_region"`): captures germline-encoded contacts
  (CDR1, CDR2, CDR2.5) that interact with the MHC molecule.
- **Single chain** (`"alpha"` or `"beta"`): useful when you suspect one
  chain dominates antigen recognition.

```{r components}
# CDR3 only (both chains, no V-region)
d_cdr3 <- tcrdist_matrix(pa_sub, "mouse", components = "cdr3")

# V-region only (CDR1 + CDR2 + CDR2.5)
d_v <- tcrdist_matrix(pa_sub, "mouse", components = "v_region")

# Components are additive: v_region + cdr3 = all
d_all <- tcrdist_matrix(pa_sub, "mouse")
stopifnot(all(abs((d_v + d_cdr3) - d_all) < 1e-10))

# Single chain
d_alpha <- tcrdist_matrix(pa_sub, "mouse", components = "alpha")
d_beta  <- tcrdist_matrix(pa_sub, "mouse", components = "beta")
stopifnot(all(abs((d_alpha + d_beta) - d_all) < 1e-10))

# Custom combination
d_custom <- tcrdist_matrix(pa_sub, "mouse", components = c("cdr3a", "vb"))
```

Available presets: `"all"` (default), `"cdr3"`, `"v_region"`, `"alpha"`,
`"beta"`. Individual terms: `"va"`, `"cdr3a"`, `"vb"`, `"cdr3b"`.

## Single-Chain Mode

Many datasets contain only beta-chain data (e.g., from bulk sequencing). All
distance functions automatically detect single-chain input and compute
distances for the available chain only:
```{r single-chain}
# Beta-only data (no alpha columns)
beta_only <- pa_sub[, c("vb", "cdr3b")]
d_beta_only <- tcrdist_matrix(beta_only, "mouse")
dim(d_beta_only)

# Result matches components="beta" on paired data
d_beta_paired <- tcrdist_matrix(pa_sub, "mouse", components = "beta")
stopifnot(all(abs(d_beta_only - d_beta_paired) < 1e-10))

# Works with all wrappers: sparse, rect, knn, radius_neighbors
knn_beta <- tcrdist_knn(beta_only, "mouse", K = 3L)
```

## Standardizing TCR Data

Use `as_tcr_df()` to convert data.frames from common tools (scRepertoire,
scirpy, dandelion, tcrdist3) to tcrdistR's canonical column names:

```{r as-tcr-df, eval=FALSE}
# Auto-detect format
tcrs <- as_tcr_df(my_data)

# Explicit format
tcrs <- as_tcr_df(my_data, format = "screpertoire")

# Custom column mapping
tcrs <- as_tcr_df(my_data, col_map = c(va = "V_alpha", cdr3a = "CDR3_alpha",
                                        vb = "V_beta",  cdr3b = "CDR3_beta"))
```

## Single-Pair Distances

For comparing individual CDR3 sequences:

```{r single-pair}
# Weighted CDR3 distance (uses BLOSUM62-derived BSD4 matrix)
weighted_cdr3_distance("CAVSLDSNYQLIW", "CALGDRATGGNNKLTF")

# Simple Hamming distance (counts mismatches at each position)
hamming_distance("CASSI", "CASSK")
```

## Diversity Metrics

Diversity metrics summarize how "spread out" a repertoire is. A repertoire
dominated by one or two clonotypes (high clonality) suggests a strong
antigen-driven expansion, while a highly diverse repertoire suggests a broad,
unstimulated sample.

```{r diversity}
# Clone counts for PA-specific TCRs
pa_counts <- pa$count

# Generalized diversity (order 2 = inverse Simpson)
# entropy: 0 = dominated by one clone, 1 = perfectly even
div <- tcr_diversity(pa_counts, order = 2)
div$entropy
div$effective_number  # "how many equally-abundant clones would give this diversity"

# Species richness: simply the number of distinct clonotypes
tcr_richness(pa_counts)

# Clonality: 0 = maximally diverse, 1 = single dominant clone
tcr_clonality(pa_counts)
```

The `effective_number` is particularly intuitive: if your repertoire has
an effective number of 50, it behaves *as if* it had 50 equally abundant
clonotypes, even though the actual distribution may be uneven.

## The TCRrep Object

For more structured workflows, wrap your data in a `TCRrep` S4 object.
By default, `TCRrep()` deduplicates identical clones (matching tcrdist3
behavior), grouping by chain columns plus `subject` if present, and summing
clone counts:

```{r tcrrep}
rep <- TCRrep(pa_sub, organism = "mouse")
nrow(pa_sub)        # before dedup
nrow(rep@clone_df)  # after dedup (within-subject duplicates merged)
rep
```

You can control deduplication with the `deduplicate` argument:

```{r tcrrep-dedup}
# Chain columns only (collapse across subjects)
rep_strict <- TCRrep(pa_sub, organism = "mouse", deduplicate = character(0))
nrow(rep_strict@clone_df)

# No deduplication
rep_raw <- TCRrep(pa_sub, organism = "mouse", deduplicate = FALSE)
nrow(rep_raw@clone_df)
```

## Choosing Parameters

Several functions take a **radius** or **threshold** parameter. Here are
practical starting points:

| Parameter | Recommended range | Notes |
|-----------|-------------------|-------|
| `tcrdist_sparse(threshold=)` | 50--200 | Use 50 for tight clusters, 200 for broad search |
| `tcrdist_radius_neighbors(radius=)` | 24--96 | 48 is a common default for paired chains |
| `neighborhood_test(radius=)` | 48--96 | Smaller radius = more specific neighborhoods |
| `find_meta_clonotypes(radius=)` | 36--72 | Balance sensitivity vs. specificity |
| `find_clumping(radii=)` | c(24, 48, 72) | Test multiple radii |
| `cluster_tcrs(k=)` | 5--20 | Start with expected number of groups |

These values assume paired alpha-beta TCRs with default weights. For
single-chain analysis, distances are roughly halved, so use proportionally
smaller thresholds.

## Performance and Scalability

All distance computations are implemented in C++ for performance.
Approximate guidance for dataset sizing:

| N (TCRs) | `tcrdist_matrix` | `tcrdist_sparse` | `tcrdist_knn` | Memory (dense) |
|----------|-------------------|-------------------|---------------|----------------|
| 1,000 | < 1 sec | < 1 sec | < 1 sec | ~8 MB |
| 5,000 | ~5 sec | ~5 sec | ~3 sec | ~200 MB |
| 10,000 | ~20 sec | ~15 sec | ~10 sec | ~800 MB |
| 50,000 | minutes | ~minutes | ~1 min | ~20 GB |

**When to use each function:**

- **< 5,000 TCRs**: `tcrdist_matrix()` is fast and convenient.
- **5,000--20,000 TCRs**: `tcrdist_sparse()` saves memory; use for
  downstream analyses that only need nearby neighbors.
- **> 20,000 TCRs**: Use `tcrdist_knn()` which avoids materializing the
  full matrix. For dimensionality reduction, use the UMAP KNN path rather
  than kernel PCA (which needs the full matrix).

## tcrdist3 Feature Map

For users migrating from Python tcrdist3:

| tcrdist3 concept | tcrdistR equivalent | Notes |
|------------------|---------------------|-------|
| `TCRrep` | `TCRrep()` S4 class | Similar dedup behavior |
| `compute_distances()` | `tcrdist_matrix()` | Per-component via `components=` |
| `compute_sparse_rect_distances()` | `tcrdist_sparse()`, `tcrdist_rect()` | |
| `nn_approach()` | `tcrdist_knn()` | |
| `TCRpublic` / `find_metaclonotypes()` | `find_meta_clonotypes()` | |
| `TCRsampler`-based clumping | `find_clumping()` | Chain-resampling background |
| Kernel PCA | `compute_tcrdist_kernel_pca()` | Linear + Gaussian kernels |
| Diversity metrics | `tcr_diversity()`, `tcr_clonality()`, etc. | |
| CDR3 motif logos | `plot_cdr3_logo()`, `plot_tcr_logo_panel()` | ggseqlogo-based |
| DB search | `match_tcrs_to_db()` | Human only |
| CD8 scoring | `make_cd8_score_table_column()` | Human only |

## Next Steps

- `vignette("tcrdistR-tcrrep-workflow")` --- End-to-end analysis with the
  TCRrep S4 object
- `vignette("tcrdistR-advanced")` --- Clustering, neighborhood tests,
  meta-clonotypes, database matching, kernel PCA, fuzzy diversity
- `vignette("tcrdistR-visualization")` --- Heatmaps, dendrograms, CDR3 logos,
  scatter plots
- `vignette("tcrdistR-comparing-repertoires")` --- Comparing repertoires
  between conditions
- `vignette("tcrdistR-glossary")` --- Glossary of technical terms
