---
title: "TCRrep Workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TCRrep Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
has_viz <- requireNamespace("ggplot2", quietly = TRUE)
has_logo <- has_viz && requireNamespace("ggseqlogo", quietly = TRUE)
has_patch <- has_viz && requireNamespace("patchwork", quietly = TRUE)
```

## Introduction

The `TCRrep` S4 object is the central data structure in tcrdistR. It holds
the clonotype table, organism, distance matrices, and downstream results in a
single object. This vignette walks through a complete analysis using TCRrep
as the backbone.

## Creating a TCRrep

Load the DASH dataset and construct a TCRrep. By default, identical clones
within the same subject are deduplicated and their counts summed:

```{r create}
library(tcrdistR)
data(dash)

rep <- TCRrep(dash, organism = "mouse", compute_distances = TRUE)
rep
```

The `compute_distances = TRUE` flag computes the full pairwise distance matrix
at construction time and stores it in the `paired_dist` slot:

```{r dist-slot}
dim(rep@paired_dist)
rep@paired_dist[1:5, 1:5]
```

## Distance Heatmap

With distances already computed, pass the `paired_dist` slot directly to the
heatmap. We use a subset for readability:

```{r heatmap, eval=has_viz, fig.height=5}
library(ggplot2)

# First 50 clones
idx <- 1:50
plot_tcrdist_heatmap(rep@paired_dist[idx, idx],
                     title = "TCRdist heatmap (first 50 clones)")
```

## Distance Distribution

```{r dist-distrib, eval=has_viz}
plot_distance_distribution(rep@paired_dist, title = "Pairwise distance distribution")
```

## Gene Usage

Gene usage plots take the clone data.frame and a column name. Access the
clonotype table via `rep@clone_df`:

```{r gene-usage, eval=has_viz}
plot_gene_usage(rep@clone_df, "va", title = "V-alpha usage")
plot_gene_usage(rep@clone_df, "vb", title = "V-beta usage")
```

## Dendrogram

The dendrogram function takes a data.frame and organism --- extract from the
TCRrep slots. We subsample for a cleaner plot:

```{r dendro, eval=has_viz, fig.height=5}
sub_idx <- sample(nrow(rep@clone_df), 100)
plot_tcrdist_dendrogram(
  rep@clone_df[sub_idx, ],
  organism = rep@organism,
  color_by = rep@clone_df$epitope[sub_idx],
  title = "TCR dendrogram colored by epitope"
)
```

## Kernel PCA and Scatter Plots

Kernel PCA projects the high-dimensional distance matrix into a few key
axes (principal components) that capture the most variation. TCRs that
cluster together in this 2D view share similar sequences --- if they also
share the same epitope label, that confirms that TCR sequence similarity
predicts antigen specificity.

Compute a kernel PCA embedding from the TCRrep:

```{r kpca}
pca <- compute_tcrdist_kernel_pca(
  rep@clone_df, organism = rep@organism,
  n_components = 50L, method = "eigen"
)
dim(pca$embeddings)
```

Visualize the first two components, colored by epitope:

```{r scatter, eval=has_viz, fig.height=5}
plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = rep@clone_df$epitope,
  title = "Kernel PCA of DASH TCRs",
  point_size = 1.5
)
```

## UMAP

UMAP provides a nonlinear 2D embedding that often separates clusters better
than linear PCA. `compute_tcrdist_umap()` supports two modes:

- **KNN path** (recommended): pass `tcr_df` + `organism`. Computes TCRdist
  K-nearest-neighbors with chain group masking, builds a fuzzy simplicial set
  graph, and runs UMAP from the precomputed KNN. This preserves the TCRdist
  metric faithfully.
- **PCA path**: pass pre-computed `pca_embeddings`. Runs standard UMAP in
  Euclidean space.

```{r umap, eval=has_viz && requireNamespace("uwot", quietly = TRUE), fig.height=5}
# KNN path: UMAP directly from TCRdist neighbors
umap <- compute_tcrdist_umap(
  rep@clone_df, organism = rep@organism, seed = 42
)
plot_tcr_scatter(
  umap$embeddings,
  color_by = rep@clone_df$epitope,
  title = "UMAP of DASH TCRs (KNN path)",
  axis_label_prefix = "UMAP",
  point_size = 1.5
)
```

The KNN path also returns the fuzzy simplicial set graph and weighted
nearest-neighbor distances:

```{r umap-knn-artifacts, eval=has_viz && requireNamespace("uwot", quietly = TRUE)}
dim(umap$knn_graph)         # N x N sparse matrix
summary(umap$nndists)       # weighted NN distances per clone
```

Alternatively, use the PCA path for a quicker approximation:

```{r umap-pca, eval=has_viz && requireNamespace("uwot", quietly = TRUE), fig.height=5}
umap_pca <- compute_tcrdist_umap(pca_embeddings = pca$embeddings, seed = 42)
plot_tcr_scatter(
  umap_pca$embeddings,
  color_by = rep@clone_df$epitope,
  title = "UMAP of DASH TCRs (PCA path)",
  axis_label_prefix = "UMAP",
  point_size = 1.5
)
```

## Single-Chain TCRrep

When working with beta-only data (e.g., from bulk sequencing), create a
TCRrep with `chains = "B"`:

```{r single-chain}
# Simulate beta-only data
beta_only <- rep@clone_df[1:50, c("vb", "cdr3b", "epitope", "subject", "count")]

rep_beta <- TCRrep(beta_only, organism = "mouse", chains = "B",
                   compute_distances = TRUE)
rep_beta
dim(rep_beta@paired_dist)

# All distance functions work with single-chain data
knn_beta <- tcrdist_knn(beta_only, "mouse", K = 3L)
```

## Clustering

Cluster TCRs and add cluster assignments to the clone table:

```{r cluster}
rep@clone_df$cluster <- cluster_tcrs(
  rep@clone_df, organism = rep@organism, k = 7
)
table(rep@clone_df$cluster)
```

Visualize clusters in PCA space:

```{r scatter-cluster, eval=has_viz, fig.height=5}
plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = as.factor(rep@clone_df$cluster),
  title = "Clusters in PCA space",
  point_size = 1.5,
  legend_title = "Cluster"
)
```

## CDR3 Sequence Logos by Epitope

Extract CDR3 sequences for a specific epitope and plot a logo:

```{r logo-pa, eval=has_logo, fig.height=3}
pa_idx <- rep@clone_df$epitope == "PA"
plot_cdr3_logo(
  rep@clone_df$cdr3b[pa_idx],
  chain = "beta", method = "bits",
  title = "CDR3-beta logo (PA epitope)"
)
```

```{r logo-np, eval=has_logo, fig.height=3}
np_idx <- rep@clone_df$epitope == "NP"
plot_cdr3_logo(
  rep@clone_df$cdr3b[np_idx],
  chain = "beta", method = "bits",
  title = "CDR3-beta logo (NP epitope)"
)
```

## Neighborhood Test

The neighborhood test asks: for each TCR, are the nearby TCRs (within a
distance radius) enriched for a particular label compared to the overall
dataset? A significant result means that antigen specificity is encoded in
the TCR sequence --- nearby TCRs tend to recognize the same epitope.

Test whether epitope labels are enriched in TCR neighborhoods:

```{r nhood}
nhood <- neighborhood_test(
  rep@clone_df, organism = rep@organism,
  variable = rep@clone_df$epitope,
  radius = 50, test = "chisq"
)
head(nhood[order(nhood$p_adjusted), c("index", "n_neighbors", "p_value", "p_adjusted")])
```

## Meta-Clonotypes

Meta-clonotypes are groups of similar TCRs found in multiple individuals.
They represent convergent immune responses --- different people independently
generate similar TCRs against the same antigen. These shared motifs can
serve as biomarkers of antigen exposure or vaccine response.

Find TCR motifs shared across subjects:

```{r meta}
meta <- find_meta_clonotypes(
  rep@clone_df, organism = rep@organism,
  radius = 48, min_nsubject = 2L,
  subject_col = "subject"
)
nrow(meta)
head(meta[, c("cdr3a", "cdr3b", "radius", "K_neighbors", "nsubject")])
```

Store back into the TCRrep:

```{r meta-store}
rep@meta_clonotypes <- meta
```

## Diversity

Diversity metrics summarize how "spread out" a repertoire is. High clonality
(close to 1) suggests antigen-driven clonal expansion; low clonality (close
to 0) suggests a broad, polyclonal sample.

```{r diversity}
div <- tcr_diversity(rep@clone_df$count, order = 2)
div$effective_number  # "equivalent number of equally-abundant clonotypes"

tcr_clonality(rep@clone_df$count)  # 0 = diverse, 1 = dominated
```

## Combined Panel

```{r panel, eval=has_patch, fig.width=10, fig.height=8}
library(patchwork)

p1 <- plot_tcrdist_heatmap(rep@paired_dist[1:40, 1:40], title = "Heatmap")
p2 <- plot_distance_distribution(rep@paired_dist, title = "Distribution")
p3 <- plot_tcr_scatter(
  pca$embeddings[, 1:2],
  color_by = rep@clone_df$epitope,
  title = "Kernel PCA", point_size = 1
)
p4 <- plot_gene_usage(rep@clone_df, "vb", title = "V-beta usage")

(p1 | p2) / (p3 | p4) +
  plot_annotation(title = "DASH Repertoire Overview")
```

## Integration with Other Tools

tcrdistR works with standard R data.frames, making it easy to integrate
with other single-cell and repertoire analysis tools:

- **Seurat / scRepertoire**: Extract TCR data from Seurat metadata using
  `as_tcr_df()`, run tcrdistR analyses, then add results (clusters,
  diversity scores) back to the Seurat object's metadata.
- **immunarch**: Use tcrdistR for distance-based analyses (neighborhood
  tests, meta-clonotypes) and immunarch for its gene usage statistics and
  tracking plots.
- **Downstream**: All tcrdistR results are standard R objects (data.frames,
  matrices, ggplot2 plots) that work with any R pipeline.

```{r seurat-example, eval=FALSE}
# Example: Seurat integration (pseudocode)
# tcr_meta <- seurat_obj@meta.data[, c("TRAV", "CDR3a", "TRBV", "CDR3b")]
# tcrs <- as_tcr_df(tcr_meta, col_map = c(va="TRAV", cdr3a="CDR3a",
#                                          vb="TRBV", cdr3b="CDR3b"))
# clusters <- cluster_tcrs(tcrs, "human", k = 10)
# seurat_obj$tcr_cluster <- clusters
```

## Session Info

```{r session-info}
sessionInfo()
```
