---
title: "Comparing TCR Repertoires"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing TCR Repertoires}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  dpi = 96
)
has_viz <- requireNamespace("ggplot2", quietly = TRUE)
has_patch <- has_viz && requireNamespace("patchwork", quietly = TRUE)
```

## Introduction

Comparing TCR repertoires between conditions is one of the most common
analysis tasks in immunology. Typical questions include:

- Do responders and non-responders share different TCR signatures?
- How does the repertoire change before and after treatment?
- Which epitope-specific responses are more clonally expanded?
- Are there public clonotypes shared across patients?

This vignette demonstrates comparison workflows using the DASH dataset,
where we treat different epitope-specific responses as different
"conditions" to compare.

## Setup

```{r setup}
library(tcrdistR)
data(dash)

# Split into two "conditions" for comparison
pa <- dash[dash$epitope == "PA", ]
np <- dash[dash$epitope == "NP", ]

cat("PA-specific TCRs:", nrow(pa), "\n")
cat("NP-specific TCRs:", nrow(np), "\n")
```

## Repertoire Overlap

**Biological question:** Do two samples share the same clonotypes?

`tcr_repertoire_overlap()` compares clone sharing between two samples
using three complementary metrics:

- **Jaccard index**: What fraction of unique clonotypes are shared?
  (presence/absence only)
- **Morisita-Horn**: Do the two samples share the same *dominant*
  clonotypes? (abundance-weighted)
- **Overlap coefficient**: What fraction of the *smaller* sample's
  clonotypes are found in the larger sample?

```{r overlap}
# Create named count vectors (names = clonotype identifiers)
pa_counts <- table(pa$cdr3b)
np_counts <- table(np$cdr3b)

ov <- tcr_repertoire_overlap(pa_counts, np_counts)
ov
```

A Jaccard of 0 means no shared CDR3-beta sequences between PA and NP
responses --- these epitopes drive completely distinct TCR populations.
This is expected for well-separated viral epitopes.

### Pairwise overlap across all epitopes

```{r overlap-matrix}
epitopes <- sort(unique(dash$epitope))
n_ep <- length(epitopes)
jaccard_mat <- matrix(0, n_ep, n_ep, dimnames = list(epitopes, epitopes))

for (i in seq_len(n_ep)) {
    counts_i <- table(dash$cdr3b[dash$epitope == epitopes[i]])
    for (j in seq_len(i)) {
        counts_j <- table(dash$cdr3b[dash$epitope == epitopes[j]])
        ov <- tcr_repertoire_overlap(counts_i, counts_j, metrics = "jaccard")
        jaccard_mat[i, j] <- ov$jaccard
        jaccard_mat[j, i] <- ov$jaccard
    }
}

round(jaccard_mat, 3)
```

Epitopes with higher overlap share more CDR3 sequences, which could
indicate cross-reactive TCRs or convergent V-gene usage.

## Comparative Diversity

**Biological question:** Which immune responses are dominated by a few
expanded clones, and which are broadly polyclonal?

A highly clonal response (clonality near 1) suggests strong
antigen-driven expansion of specific T cell clones. A diverse response
(clonality near 0) suggests either a polyclonal response or insufficient
stimulation.

```{r diversity-comparison}
diversity_table <- do.call(rbind, lapply(epitopes, function(ep) {
    counts <- dash$count[dash$epitope == ep]
    div <- tcr_diversity(counts, order = 2, ci = FALSE)
    data.frame(
        epitope = ep,
        n_clones = tcr_richness(counts),
        clonality = round(tcr_clonality(counts), 3),
        shannon = round(tcr_shannon_entropy(counts), 3),
        simpson = round(div$entropy, 3),
        effective_n = round(div$effective_number, 1),
        gini = round(tcr_gini(counts), 3),
        stringsAsFactors = FALSE
    )
}))

diversity_table
```

**Interpreting the results:**

- **n_clones** (richness): How many distinct clonotypes were detected.
- **clonality**: 0 = maximally diverse, 1 = single dominant clone.
  Epitopes with high clonality have strong clonal expansions.
- **effective_n**: "How many equally-abundant clonotypes would give this
  diversity?" Lower effective_n relative to n_clones means the repertoire
  is dominated by a few clones.
- **gini**: Inequality in clone sizes. High Gini means a few clones
  dominate; low Gini means clones are evenly distributed.

## Visualizing Repertoire Differences

### Gene usage comparison

**Biological question:** Do different epitope responses use different
V-genes?

```{r gene-usage-compare, eval=has_patch, fig.width=10, fig.height=4}
library(ggplot2)
library(patchwork)

p1 <- plot_gene_usage(pa, "vb", title = "V-beta usage (PA)")
p2 <- plot_gene_usage(np, "vb", title = "V-beta usage (NP)")
p1 | p2
```

Differences in V-gene usage between conditions can indicate germline-encoded
contributions to antigen recognition. A V-gene enriched in one condition
may encode CDR1/CDR2 loops that contact the target MHC-peptide complex.

### CDR3 length distributions

```{r cdr3-length, eval=has_patch, fig.width=10, fig.height=4}
p1 <- plot_cdr3_length(pa, chain = "beta", title = "CDR3-beta length (PA)")
p2 <- plot_cdr3_length(np, chain = "beta", title = "CDR3-beta length (NP)")
p1 | p2
```

CDR3 length differences can reflect constraints on peptide binding ---
some epitopes require longer or shorter CDR3 loops for optimal contact.

### Scatter plots with faceting and highlighting

Visualize the combined repertoire in 2D, colored and faceted by epitope:

```{r scatter-facet, eval=has_viz, fig.height=5, fig.width=8}
# Combine PA and NP for joint embedding
combined <- rbind(pa[1:100, ], np[1:100, ])
pca <- compute_tcrdist_kernel_pca(combined, "mouse", n_components = 5L)

plot_tcr_scatter(
    pca$embeddings[, 1:2],
    color_by = combined$epitope,
    title = "PA vs NP in PCA space",
    point_size = 1.5
)
```

Highlight one epitope against the background of the full repertoire:

```{r scatter-highlight, eval=has_viz, fig.height=5}
plot_tcr_scatter(
    pca$embeddings[, 1:2],
    metadata = combined,
    color_by = "epitope",
    highlight = list(epitope = "PA"),
    title = "PA clones highlighted",
    point_size = 1.5
)
```

## Shared Clonotype Identification

**Biological question:** Are there public TCR sequences shared across
multiple individuals?

Public clonotypes --- identical or near-identical TCR sequences found in
different people --- suggest convergent immune responses driven by common
antigens. These can serve as biomarkers of infection or vaccination.

```{r public-clonotypes}
# For PA-specific TCRs, count how many subjects share each CDR3-beta
pa_by_subject <- split(pa$cdr3b, pa$subject)
all_cdr3b <- unique(pa$cdr3b)

# Count subjects per clonotype
subject_counts <- sapply(all_cdr3b, function(seq) {
    sum(sapply(pa_by_subject, function(seqs) seq %in% seqs))
})

# Most public clonotypes
public <- sort(subject_counts, decreasing = TRUE)
head(public, 10)

cat("\nClonotypes found in >= 3 subjects:",
    sum(public >= 3), "of", length(public), "\n")
```

For distance-based public motifs (allowing near-matches across
individuals), use `find_meta_clonotypes()`:

```{r meta-public}
meta <- find_meta_clonotypes(
    pa, organism = "mouse",
    radius = 48, min_nsubject = 3L,
    subject_col = "subject"
)
cat("Meta-clonotypes shared across >= 3 subjects:", nrow(meta), "\n")
if (nrow(meta) > 0) {
    head(meta[, c("cdr3a", "cdr3b", "K_neighbors", "nsubject")])
}
```

## Distance-Based Comparison

**Biological question:** How structurally different are two repertoires
in TCR sequence space?

Use `tcrdist_join()` to find TCRs in one sample that are similar to TCRs
in another:

```{r distance-join}
# Find PA TCRs within distance 100 of NP TCRs
pa_sub <- pa[1:50, ]
np_sub <- np[1:50, ]
cross_matches <- tcrdist_join(pa_sub, np_sub, organism = "mouse",
                              radius = 100, max_n = 3)
cat("Cross-epitope TCR pairs within radius 100:", nrow(cross_matches), "\n")
```

Few or no cross-matches at a tight radius confirms that the two epitope
responses use structurally distinct TCR repertoires.

## Session Info

```{r session-info}
sessionInfo()
```
