---
title: "Getting Started with tcrdistR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with tcrdistR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

**tcrdistR** computes pairwise distances between T-cell receptors (TCRs) using
the TCRdist metric. Each TCR is represented by its V-gene and CDR3 amino acid
sequence for both alpha and beta chains. The distance incorporates V-region
similarity (via BLOSUM62-derived substitution matrices) and CDR3 sequence
alignment with variable gap positioning matching tcrdist3.

All distance computations are implemented in C++ for high performance.

## Installation

```{r install, eval=FALSE}
# install.packages("devtools")
devtools::install_github("shihanli92/tcrdistR")
```

## The DASH Dataset

tcrdistR ships with the **DASH** dataset (Dash et al., 2017) --- 1924 paired
alpha-beta mouse TCRs responding to 7 viral epitopes, collected from 78
subjects. We use it throughout these vignettes.

```{r load-data}
library(tcrdistR)
data(dash)

dim(dash)
colnames(dash)
table(dash$epitope)
```

The required columns for distance computation are `va`, `cdr3a`, `vb`, and
`cdr3b`. Additional columns like `epitope` and `subject` are carried along for
downstream analysis.

## Reading TCR Data from Files

tcrdistR can read TCR data from multiple common formats. All readers produce
a data.frame with standardized column names (`va`, `cdr3a`, `vb`, `cdr3b`).

```{r read-data, eval=FALSE}
# Auto-detect format from column naming conventions
tcrs <- read_tcr_table("my_data.tsv")

# Format-specific readers
tcrs <- read_10x("filtered_contig_annotations.csv")
tcrs <- read_airr("airr_rearrangements.tsv")
tcrs <- read_adaptive("immunoseq_export.tsv")
```

## Computing the Distance Matrix

The core function `tcrdist_matrix()` computes a dense N x N pairwise distance
matrix. We use a subset of 50 PA-specific TCRs for speed:

```{r dist-matrix}
pa <- dash[dash$epitope == "PA", ]
pa_sub <- pa[1:50, ]

dist_mat <- tcrdist_matrix(pa_sub, organism = "mouse")
dim(dist_mat)
dist_mat[1:5, 1:5]
```

Each entry is an integer TCRdist value. Identical TCRs have distance 0;
typical distances range from 0 to ~400.

## Sparse Distances

For large datasets, storing the full N x N matrix is impractical. Use
`tcrdist_sparse()` to only store distances below a threshold:

```{r sparse}
sparse_mat <- tcrdist_sparse(pa_sub, organism = "mouse", threshold = 100)
class(sparse_mat)
```

The result is a `Matrix::dgCMatrix`. Non-stored entries represent distances
above the threshold.

## Rectangular Distances

Compute distances between two different sets of TCRs (query vs reference):

```{r rect}
query <- pa_sub[1:5, ]
reference <- pa_sub[6:50, ]
rect_mat <- tcrdist_rect(query, reference, organism = "mouse")
dim(rect_mat)
```

## K-Nearest Neighbors

Find the K closest TCRs for each input. This computes distances and extracts
neighbors in a single pass without materializing the full N x N matrix:

```{r knn}
knn <- tcrdist_knn(pa_sub, organism = "mouse", K = 5L)
dim(knn$knn_indices)    # N x K
dim(knn$knn_distances)  # N x K

# Nearest neighbor of TCR 1
knn$knn_indices[1, ]
knn$knn_distances[1, ]
```

## Radius-Based Neighbors

Find all neighbors within a distance threshold:

```{r radius}
neighbors <- tcrdist_radius_neighbors(pa_sub, organism = "mouse", radius = 50)
length(neighbors)  # one entry per TCR

# First TCR's neighbors
neighbors[[1]]$indices
neighbors[[1]]$distances
```

## Single-Pair Distances

For comparing individual CDR3 sequences:

```{r single-pair}
# Weighted CDR3 distance (uses BLOSUM62-derived BSD4 matrix)
weighted_cdr3_distance("CAVSLDSNYQLIW", "CALGDRATGGNNKLTF")

# Simple Hamming distance (counts mismatches at each position)
hamming_distance("CASSI", "CASSK")
```

## Diversity Metrics

Repertoire diversity functions work on clone count vectors:

```{r diversity}
# Clone counts for PA-specific TCRs
pa_counts <- pa$count

# Generalized diversity (order 2 = inverse Simpson)
div <- tcr_diversity(pa_counts, order = 2)
div$entropy
div$effective_number

# Species richness
tcr_richness(pa_counts)

# Clonality (1 - normalized Shannon entropy)
tcr_clonality(pa_counts)
```

## The TCRrep Object

For more structured workflows, wrap your data in a `TCRrep` S4 object:

```{r tcrrep}
rep <- TCRrep(pa_sub, organism = "mouse", chains = "AB")
rep
```

## Next Steps

- `vignette("tcrdistR-advanced")` --- Clustering, neighborhood tests,
  meta-clonotypes, database matching, kernel PCA, fuzzy diversity
- `vignette("tcrdistR-visualization")` --- Heatmaps, dendrograms, CDR3 logos,
  scatter plots
